DeepSphere: Efficient spherical Convolutional Neural Network with HEALPix sampling for cosmological applications

### 2019-03-04 rebuttal.tex

\subsection{}
\begin{mdframed}[style=comment]
Page 4: "There is no known point set that achieves the analogue of uniform sampling in Euclidean space and allows exact and invertible discrete spherical harmonic decompositions of arbitrary but band-limited functions." This is ambiguous. There exists multiple schemes for performing exact SHT on band-limited signals. Those sampling are more efficient than healpix - often in terms of number of pixels, and sometimes in terms of performance of the spherical harmonic transforms - and more accurate since the transform is exact (at least numerically - so typical errors are machine precision). Healpix transforms are only exact to 1e-7 when using iterative schemes, and that depends on Nside and lmax. Please clarify this paragraph. Obviously healpix is very advantageous for the pooling and other operations that take advantage of its hierarchical nature. This is well discussed in the paper - in particular how the fast graph algorithms could accommodate other sampling schemes.
\end{mdframed}

That is right.
Thanks for pointing it out.
The confusion was due to our ignorance.
We removed the part about exact SHTs from the paragraph as we don't need to make such a point.
%\mdeff{That sentence mostly came from \url{https://healpix.jpl.nasa.gov/html/intronode2.htm}.}
%\mdeff{uniform sampling == same distance between all neighboring pixels}

\subsection{}
\begin{mdframed}[style=comment]
"Our graph is constructed as an approximation of sphere S2, a 2D manifold embedded in R3, Indeed, [33] showed that the graph Laplacian converges to the Laplace-Beltrami when the number of pixels goes to infinity. While our construction will not exactly respect the setting defined by [33], we observe empirically strong evidence of convergence." This paragraph is unclear. Could you please add a few words to clarify what the approach of 33 is?
\end{mdframed}

We tried to clarify the paragraph in the following manner:
\begin{mdframed}[style=manuscript]
Our graph is constructed as an approximation of the sphere $S^2$, a 2D manifold embedded in $\mathbb{R}^3$.
Indeed, \cite{belkin2007convergence} showed that the graph Laplacian converges to the Laplace-Beltrami when the number of pixels goes to infinity providing uniform sampling of the manifold and a fully connected graph built with exponentially decaying weights.
While our construction does not exactly respect their setting (the sampling is deterministic and the graph is not fully connected), we empirically observe a strong correspondence between the eigenmodes of both Laplacians (see Appendix A).
\end{mdframed}

### 2019-03-04 rebuttal.tex

- What does "this" refer to?
\end{mdframed}

We precised it.

\begin{mdframed}[style=comment]
In a neural network setting, the fact that the graph convolution is not exactly equivariant to rotation is probably compensated by the fact that convolution kernels are learned.

- I think it might be another one of the most important points in the paper which allows you to justify why not being exact (like a spherical harmonic transform) is acceptable. If I were you I'd move this into the main body of the paper.
\end{mdframed}

Thanks for the suggestion.
We adopted it.

% \todo{To check: @all}
% \mdeff{I added a similar sentence in the related work section.}

% We agree it is an important point, especially when one discusses rotation equivariance.
% However, since almost all of this discussion is done in the appendix, it is complicated to bring this point to the main body without moving a significant part of the appendix as well.

@@

\begin{mdframed}[style=comment]
We then plot the result with a gnomonic projection.

- Can you add in Figure C.16 here. "We then plot the result with a gnomonic projection, as shown in Figure C.16". Also, why do you choose a gnomonic projection? In fact, with plots C.16 and C.17 I would make them a single plot (as in Figure. 10),
\end{mdframed}

Inspired by the reviewer comment, we have merged the three plots in a single figure and adapted the text accordingly.

@@

\begin{mdframed}[style=comment]
We plot the section of the convoluted Kroneker.

- It should be "convolved" and probably stick with "Kroneker $\delta$". Also, why are you plotting this? What does it show? Can you go into a bit more detail?
\end{mdframed}

This plot shows the section of the filter.
We believe that it might be easier to read than the gnomonic projection, as the filter is isotropic.
We have modified the text.

@@

- I think this should be "converges such that" rather than "converged and". And rather than "help" I think "improve convergence further" is more descriptive.
\end{mdframed}

We changed the sentence to:
\begin{mdframed}[style=manuscript]
The error clearly converged such that adding more samples will not improve the classifier.
\end{mdframed}

\begin{mdframed}[style=comment]
The validation error might be slightly below the training error as it is used to select hyper-parameters.

- I don't understand this statement. It could be that there is accidentally more correlation in the validation set, but I'm not sure what you mean by "it is used to select hyper-parameters" or why that would explain why there is less error in the validation set. Can you think about this further (and perhaps check for the amount of information in that data set compared to the training set)?
\end{mdframed}

% We changed the sentence as:
% \begin{mdframed}[style=manuscript]
% The validation error might be slightly below the training error as the validation set is used to select hyper-parameters.
% \end{mdframed}

For linear SVM, there is one hyper-parameter to tune. As we select the hyperparameter with the lowest validation error, it is possible that the validation error is lower than the training error.

@@

This has been corrected.

\subsection{}

\begin{mdframed}[style=comment]

Finally, after all these little points, I just want to congratulate the authors again on a great paper. It was a pleasure reading it. I will be sure to be using this method in the near future!
\end{mdframed}

We would like to thank the reviewer again for his exceptionally detailed review.
We really appreciate his effort and the time he invested in his answers.
He is welcome to contact us again for additional information or for a discussion post-publication.

### 2019-03-04 rebuttal.tex

% \mdeff{Agreed that we should do at least one thing they asked for.
% \begin{itemize}
% 	\item Testing on the 1/12 of the sphere is not what he asked for. On one hand, a global projection would exhibit more distortion. On the other, we never use the whole sphere.
% 	\item Keep in mind that there is no orientation that is consistent for the 12 base faces (i.e., where is up?).
% 	\item Agreed that the performance should be very similar. Both are doing local computations. And locally (3 to 5 pixels across), the sphere is isomorphic to an Euclidean plane.
% 	\item The only difference, in my opinion, would be due to the anisotropic filters the 2D CNN has access to.
% 	\item Not sure we should include this result in the main paper. Hard to justify in the story, especially as a comparison with alternative spherical CNNs is missing. Let's see the results.
% \end{itemize}
% }

\subsection{}

\begin{mdframed}[style=comment]
Similarly, I think it would useful to check how well the DeepSphere CNN does against the FCN when the input data is rotated randomly during training - I imagine that the results should match or probably exceed the FCN results and I think this would be a neat little test which doesn't involve needing extra data, just more augmentation.
\end{mdframed}

% That comparison is not warranted.
% or justified. pointless is maybe too aggressive
We do not believe this comparison is useful because of the following theoretical reason.
There are two ways to deal with symmetries one wants to be invariant to: (i) build them into the architecture of the model, or (ii) augment the dataset such that a (more general) model learns them ``brute-force''.
With respect to rotation invariance, the FCN architecture is of the first kind, while the CNN is of the second kind.
Assuming the task is truly invariant to said symmetry,\footnote{As argued in the introduction, cosmological maps are rotation invariant.} the performance of the first kind of models would be exactly the same as without augmentation, while the second kind would only catch up after seeing enough data to learn the invariance.
% That is a waste of resources which is only justified if it's not possible to back the desired invariance in the architecture.
Augmentation is only justified if it's not possible to back the desired invariance in the architecture.
% The reviewer himself makes that point in comment 2.8: ``building a NN to suit your purposes''.
We updated the manuscript to more precisely reflect this argumentation.

%\TK{Seems a bit strong to me.. maybe we would write it in a softer language?}
%\nati{Now it is OK. But is it still strong enough. We should only point reviewer incoherence if necessary.}

@@

Thank you very much for your comments. We addressed them to the best we could.

\subsection{}
\begin{mdframed}[style=comment]
While our construction will not exactly respect the setting defined by [33], we observe empirically strong evidence of convergence.

- Could you add a plot comparing the graph Laplacian as a function of number of pixels and the Laplace-Beltrami operator so the reader can explicitly see the convergence?
\end{mdframed}

Good point.
To experimentally show convergence, we would need to build a series of graph Laplacians that converges to the Laplace-Beltrami operator.
This is complicated and requires to solve the theoretical problem first.
Hence, we built one Laplacian and observed some similarities between the two operators instead.
We agree that our wording was unfortunate and we have modified the manuscript to mean that we empirically observed a strong correspondence rather than a proper convergence.
The convergence (or even exact correspondence) is a problem we are currently investigating and plan to release as a separate contribution.
See also comment 1.4 and our answer there.

%\todo{the new figure A.12 is maybe a weak proof of convergence}

%\mdeff{@nati, what do you think?} \nati{I like it. I totally agree with the new wording.}

@@

Think of the MNIST digits, cells in biomedical imaging, faces on portraits, etc.
Most photographers prefer the subject of interest to be centered.
Then, what's important is not that the informative part is only in the center, but that a fully connected layer allows a network to focus its attention on the domain.
With average pooling, all pixels are treated equally.

\subsection{}
\begin{mdframed}[style=comment]
Than can be seen as intrinsic data augmentation, as a CNN would need to see many rotated versions of the same data to learn the invariance.

- I think you mean "This" rather than "Than". I think, however, you are making a more important point here than you realise. It is not really "intrinsic data augmentation" but rather building a NN to suit your purposes. This should be done by the entire community as, at the moment, too many people blindly build their networks without considering their data and then struggle to get the best results out. The symmetries of the data should be the first thing to influence architecture of a network, exactly as you have reasoned. You might want to add a statement about how you know what your symmetries are and thus, you know how to build a well reasoned network which is more likely to work (and that other should do the same).
\end{mdframed}

Thanks for pointing out the typo, we fixed it.
We do realize it's an important point, and agree that some people apply architectures to a different problem they have been devised for without much thought.
A large chunk of our paper is actually about using an appropriate NN architecture.
The point you suggest us adding is actually made in the Experiments section, where we describe the data and task, and motivate the architecture choice.
We wanted the Method section to be more generic.
% Note that the task plays a role as well (i.e., two different tasks on the same data might benefit from different architectures).
% \mdeff{that's in contradiction with his comment 2.3 where he asks us to do data augmentation with the CNN}

@@

- I don't understand what the difference is between the convolution and the "non-linearity". I think this sentence either needs explaining further. Is the process $\mathcal{O}(N_{pix})$ or $\mathcal{O}(N^{3/2}_{pix})$?
\end{mdframed}

In \cite{kondor2018clebsch}, the convolution is performed in the spectral (harmonic) domain with a simple Hadamard product.
Instead of recomposing the signal and applying a traditional non-linearity such as the ReLU in the pixel domain, \cite{kondor2018clebsch} proposes to apply a Clebsch–Gordan product in the spectral domain.
(As a non-linear operation preserving rotation equivariance, it is considered as the non-linearity of the NN.)
Their convolution costs $\mathcal{O}(N_{pix})$, and their non-linearity costs $\mathcal{O}(N^{3/2}_{pix})$.
Stacking those layers result in a NN of overall $\mathcal{O}(N^{3/2}_{pix})$ time complexity.
We clarified the sentence.

\subsection{}
\begin{mdframed}[style=comment]
When mentioning that you are 10 to 20 times faster than the SHT it should be noted that libsharp has amazing distributed mpi functionality which helps, although of course your method is genuinely lower order computationally (and can be distributed simply on a GPU).
\end{mdframed}

True.
In the experiment, both were compared on a single core.
They might, however, scale differently given one can be distributed through MPI and the other on a GPU.
The manuscript was updated.

### 2019-03-04 scnn.tex

Rotation equivariance was addressed by leveraging the convolution associated to the 3D rotation group SO(3), with applications to atomization energy regression and 3D model classification, alignment and retrieval \citep{cohen2018sphericalcnn,esteves2017sphericalcnn}.
The resulting convolution is performed by (i) a spherical harmonic transform (SHT), i.e., a projection on the spherical harmonics, (ii) a  multiplication in the spectral domain, and (iii) an inverse SHT.
Note the similarity with the naive graph convolution defined in \eqnref{graph_convolution_fourier}.
Likewise, the computational cost of a convolution is dominated by the two SHTs, and a naive implementation of the SHT costs $\bO(N_{pix}^2)$ operations.
Accelerated schemes however exist for some sampling sets \citep[see][for examples]{mohlenkamp1999fast, rokhlin2006fast, reinecke2013libsharp}.
The convolutions remain nevertheless expensive, limiting the practical use of this approach.
For example with HEALPix, which was designed to have a fast SHT by being iso-latitude, the computational cost of the SHT is $\bO(N_{pix}^{3/2}) = \bO(N_{side}^3) = \bO(\ell_{max}^3)$, where $\ell_{max}$ is the largest angular frequency \citep{gorski2005healpix, reinecke2013libsharp}.\footnote{All pixels are placed on $N_{ring} = 4N_{side}-1 = \bO\left(\sqrt{N_{pix}}\right)$ rings of constant latitude.
Each ring has $\bO\left(\sqrt{N_{pix}}\right)$ pixels.
Thanks to this iso-latitude property, the SHT is computed using recurrence relations for Legendre polynomials on co-latitude and fast Fourier transforms (FFTs) on longitude.
The computational cost is thus $\sqrt{N_{pix}}$ FFTs for a total cost of $\bO\left( N_{pix} \log \sqrt{N_{pix}} \right)$, plus $\sqrt{N_{pix}}$ matrix multiplications of size $\sqrt{N_{pix}}$ for a total cost of $\bO\left(N_{pix}^{3/2}\right)$ operations.}
While the Clebsh-Gordan transform can be leveraged to avoid SHTs between layers (it does a non-linear transformation in the spectral domain), the transform itself costs $\bO(N_{pix}^{3/2})$, for no reduction of overall complexity \citep{kondor2018clebsch}.
This work is another indication that there might be a $\Omega(N_{pix}^{3/2})$ computational lower bound for the proper treatment of rotation equivariance with the spherical harmonics.
In comparison, DeepSphere scales as $\bO(N_{pix})$ (see \secref{efficient_convolution}), at the expense of not being exactly equivariant (see \ref{sec:comparison_spherical_harmonics}).
Being a mathematically well-defined rotation equivariant network is the main advantage of methods based on spherical harmonics.
See \citep{kondor2018equivariance} for a rigorous treatment of convolution and equivariance in neural networks.

@@

Let us consider the heat diffusion problem
\begin{equation}
  \tau \L \b{f}(t) = - \partial_t \b{f}(t),
  \label{eqn:heat_equation}
\end{equation}
where $\b{f}: \R_+ \rightarrow \R^{N_{pix}}$. Given the initial condition
$\b{f}(0)$, the solution of~\eqnref{heat_equation} can be expressed as
\begin{equation*}
  \b{f}(t) = e^{-\L \tau t} \b{f}(0) = \U e^{-\bLambda \tau t} \U\trans \g{f}(0) = K_t(\L) \b{f}(0),
\end{equation*}
which is, by definition, the convolution of the signal $\b{f}(0)$ with the kernel $K_t(\lambda)=e^{-\tau t \lambda}$. Since the kernel $K_t$ is applied to the graph eigenvalues $\bLambda$, which can be interpreted as squared frequencies, it can also be considered as a generalization of the Gaussian kernel on the sphere. \figref{gaussian_filters_comparizon} shows the effect of the convolution by diffusing a unit of heat for $\tau=1$ at various times $t$. By comparing the graph convolution with the spherical symmetric Gaussian smoothing, we observe that both techniques lead to similar results (see \figref{gaussian_filters_comparizon}). While the graph convolution remains different from the spherical convolution, this small experiment shows that, providing the correct parameters, the graph convolution can approximate the spherical convolution. For additional insights about the graph kernel $K_t$, we refer to the \figref{gaussian_filters_visualization}.
% and~\ref{sec:filter_visualization}.

@@

\begin{figure}[ht!]
\centering
\includegraphics[width=\linewidth]{gaussian_filters_spectral}\\
\vspace{0.2in}
\includegraphics[width=\linewidth]{gaussian_filters_gnomonic}\\
\vspace{0.2in}
\hspace{0.3cm}\includegraphics[width=0.95\linewidth]{gaussian_filters_section}
\caption{\nati{Visualization of the convolution kernel $K_t(x)=e^{-\tau t x}$. (top) Spectral domain. (middle) Cross-section of the kernel through the pixels along the equator. (bottom) Gnomonic projection.}}
\label{fig:gaussian_filters_visualization}
\end{figure}

In \figref{gaussian_filters_visualization}, we further leverage this example to present the three visualizations of a HEALPix graph convolution kernel.
We remember that the convolution kernel resulting from the heat diffusion is defined as $K_t(x)=e^{-\tau t x}$. For our (top) visualization, we simply plot this function, i.e., we evaluate $K_t$
at the graph eigenvalues $\text{diag}(\bLambda)$. Note that, the lower eigenvalues $\lambda$ correspond to the lower frequencies, i.e., the spectral mode with less variation.
As a second visualization (middle), we observe the kernel directly on the sphere. As the kernel $K_t$ is defined in the graph spectral domain, we convolve a Kronecker $\delta$ to obtain a representation on the sphere. We then plot the result with a gnomonic projection.
\nati{For our third visualization (bottom), our goal is to plot the section of a filter. This visualization might be convenient as we theoretically have spherical filters (up to the irregularities of the sampling). The idea is to plot the section of the convolved Kronecker $\delta$.}
To do so, we convolve a Kronecker $\delta$ on the vertices along the equator and use the equator for our section.
Note that, because of the small irregularities in the HEALPix sampling, the second and the third methods are subject to small variations depending on the chosen position of the Kronecker $\delta$.

@@

\begin{figure}[ht!]
	\centering
	\includegraphics[width=\linewidth]{hist_error_order2_noise1_5}
	\caption{Error w.r.t.\ the number of training samples for histogram features and the linear SVM (setting: order $o=2$, relative noise level 1.5). The error clearly converged such that adding more samples will not improve the classifier. The validation error might be slightly below the training error as the validation set is used to select hyper-parameters.}
	\label{fig:hist_error_evolution}
\end{figure}

>>>>

Rotation equivariance was addressed by leveraging the convolution associated to the 3D rotation group SO(3), with applications to atomization energy regression and 3D model classification, alignment and retrieval \citep{cohen2018sphericalcnn,esteves2017sphericalcnn}.
The resulting convolution is performed by (i) a spherical harmonic transform (SHT), i.e., a projection on the spherical harmonics, (ii) a  multiplication in the spectral domain, and (iii) an inverse SHT.
Note the similarity with the naive graph convolution defined in \eqnref{graph_convolution_fourier}.
Likewise, the computational cost of a convolution is dominated by the two SHTs, and a naive implementation of the SHT costs $\bO(N_{pix}^2)$ operations.
Accelerated schemes however exist for some sampling sets \citep[see][for examples]{mohlenkamp1999fast, rokhlin2006fast, reinecke2013libsharp}.
The convolutions remain nevertheless expensive, limiting the practical use of this approach.
For example with HEALPix, which was designed to have a fast SHT by being iso-latitude, the computational cost of the SHT is $\bO(N_{pix}^{3/2}) = \bO(N_{side}^3) = \bO(\ell_{max}^3)$, where $\ell_{max}$ is the largest angular frequency \citep{gorski2005healpix, reinecke2013libsharp}.\footnote{All pixels are placed on $N_{ring} = 4N_{side}-1 = \bO\left(\sqrt{N_{pix}}\right)$ rings of constant latitude.
Each ring has $\bO\left(\sqrt{N_{pix}}\right)$ pixels.
Thanks to this iso-latitude property, the SHT is computed using recurrence relations for Legendre polynomials on co-latitude and fast Fourier transforms (FFTs) on longitude.
The computational cost is thus $\sqrt{N_{pix}}$ FFTs for a total cost of $\bO\left( N_{pix} \log \sqrt{N_{pix}} \right)$, plus $\sqrt{N_{pix}}$ matrix multiplications of size $\sqrt{N_{pix}}$ for a total cost of $\bO\left(N_{pix}^{3/2}\right)$ operations.}
While the Clebsh-Gordan transform can be leveraged to avoid SHTs between layers (it does a non-linear transformation in the spectral domain), the transform itself costs $\bO(N_{pix}^{3/2})$, for no reduction of overall complexity \citep{kondor2018clebsch}.
This work is another indication that there might be a $\Omega(N_{pix}^{3/2})$ computational lower bound for the proper treatment of rotation equivariance with the spherical harmonics.
In comparison, DeepSphere scales as $\bO(N_{pix})$ (see \secref{efficient_convolution}), at the expense of not being exactly equivariant (see \ref{sec:comparison_spherical_harmonics}).
Being a mathematically well-defined rotation equivariant network is the main advantage of methods based on spherical harmonics, though the fact that convolution kernels are learned probably compensates for inexact equivariance.
See \citep{kondor2018equivariance} for a rigorous treatment of convolution and equivariance in neural networks.

@@

Let us consider the heat diffusion problem
\begin{equation}
  \tau \L \b{f}(t) = - \partial_t \b{f}(t),
  \label{eqn:heat_equation}
\end{equation}
where $\b{f}: \R_+ \rightarrow \R^{N_{pix}}$. Given the initial condition
$\b{f}(0)$, the solution of~\eqnref{heat_equation} can be expressed as
\begin{equation*}
  \b{f}(t) = e^{-\L \tau t} \b{f}(0) = \U e^{-\bLambda \tau t} \U\trans \g{f}(0) = K_t(\L) \b{f}(0),
\end{equation*}
which is, by definition, the convolution of the signal $\b{f}(0)$ with the kernel $K_t(\lambda)=e^{-\tau t \lambda}$. Since the kernel $K_t$ is applied to the graph eigenvalues $\bLambda$, which can be interpreted as squared frequencies, it can also be considered as a generalization of the Gaussian kernel on the sphere. \figref{gaussian_filters_comparizon} shows the effect of the convolution by diffusing a unit of heat for $\tau=1$ at various times $t$. By comparing the graph convolution with the spherical symmetric Gaussian smoothing, we observe that both techniques lead to similar results (see \figref{gaussian_filters_comparizon}). While the graph convolution remains different from the spherical convolution, this small experiment shows that, providing the correct parameters, the graph convolution can approximate the spherical convolution.
\figref{gaussian_filters_visualization} should give more insights about the graph kernel $K_t$.

@@

\begin{figure}[t]
	\centering
	\includegraphics[width=\linewidth]{gaussian_filters_spectral} \\
	\vspace{0.2in}
	\includegraphics[width=\linewidth]{gaussian_filters_gnomonic} \\
	\vspace{0.2in}
	\includegraphics[width=\linewidth]{gaussian_filters_section}
	\caption{Visualization of the convolution kernel $K_t(x)=e^{-\tau t x}$. Top: spectral domain. Middle: gnomonic projection. Bottom: cross-section along the equator.}
	\label{fig:gaussian_filters_visualization}
\end{figure}

We further leverage this example to present three visualizations of a convolution kernel.
The first visualization (\figref{gaussian_filters_visualization}, top) is a plot of the kernel function, i.e., it shows $K_t$ evaluated at the graph eigenvalues $\text{diag}(\bLambda)$.
% and~\ref{sec:filter_visualization}.
Note that small eigenvalues $\lambda$ correspond to small frequencies, i.e., to spectral modes of low variation.
The second visualization (\figref{gaussian_filters_visualization}, middle) shows the kernel localized on the sphere (gnomonic projection).
As the kernel $K_t$ is defined in the graph spectral domain, it was convolved with a Kronecker $\delta$ to obtain a representation on the sphere (see \secref{graph_convolution}).
The third visualization (\figref{gaussian_filters_visualization}, bottom) plots the section of the filter.
This visualization might be easier to read as the filters are isotropic (up to the irregularities of the sampling).
The plot is obtained by convolving a Kronecker $\delta$ on the vertices along the equator.
Note that, because of the small irregularities in the HEALPix sampling, the second and third methods are subject to small variations depending on the chosen position of the Kronecker $\delta$.

@@

\begin{figure}[ht!]
	\centering
	\includegraphics[width=\linewidth]{hist_error_order2_noise1_5}
	\caption{Error w.r.t.\ the number of training samples for histogram features and the linear SVM (setting: order $o=2$, relative noise level 1.5). The error clearly converged such that adding more samples will not improve the classifier. The validation error might be slightly below the training error as it was used to select the hyper-parameters.}
	\label{fig:hist_error_evolution}
\end{figure}
